Manuscript_Github_Pchabaudi_pir_TPMHeatmaps

### 2021-04-21 Manuscript_pirAnalysis_Github/Pchabaudi24h/Manuscript_Github_Pchabaudi_pir_TPMHeatmaps.Rmd
Chapl/Aapl heatmap start

cir_clade_info <- read_delim('../standard_files/Pchabaudi_pir_clades.txt',
            delim = '\t')
```

@@

@@

cir_housekeeping_id <- c(cir_id, 
                         'PCHAS_0721000',
                         'PCHAS_0831300',
                         'PCHAS_0931000')
```

```{r}
# Get order of pir genes by highest expressed genes in each sample.

combined_data_pir <- twentyfour_avgtpm$RMT$tpm %>% 
  filter(Geneid %in% cir_housekeeping_id) %>% 
  column_to_rownames(var = 'Geneid') %>% 
  filter_all(
      all_vars(. > 1)
  )

@@

# combined_data_pir_ht <- twentyfour_avgtpm$RMT$tpm %>% 
#   column_to_rownames(var = 'Geneid')
# 
# #Get the times instead of the stage names.
# colnames(combined_data_pir_ht) <- str_extract(colnames(combined_data_pir_ht),
#                                               pattern = '[[:digit:]]{2}h')
# 
# combined_data_pir_ht <- combined_data_pir_ht[,order(colnames(combined_data_pir_ht)),
#                                              drop = FALSE]

@@

## Figure 4 and supplementary figure 7

_pir_ genes selected by being expressed above 1 TPM in all samples.

```{r pirs_side-by-side_figure4}

cir_plot <- twentyfour_rMT %>% 
  filter(Geneid %in% cir_housekeeping_id) %>%
  melt(id.vars = 'Geneid', 
       variable.name = 'sample', 
       value.name = 'tpm') %>% 
  mutate(
    stage = gsub(sample, pattern = '_[[:digit:]]$', replacement = ''),
    diff_count = str_split_fixed(sample, pattern = '_', n = 4)[,2],
    time = str_split_fixed(sample, pattern = '_', n = 3)[,1],
    subfam = cir_clade_info$clade[match(Geneid, cir_clade_info$Geneid)]
  ) %>% 
  mutate(subfam = ifelse(is.na(subfam), 'ref', subfam)) %>% 
  filter(!(time %in% c('sch.culture', 
                       'Unknown'))) %>% 
  mutate(subfam = str_replace(subfam, 
                              'ancestral',
                              'a')) %>% 
  mutate(subfam =
           forcats::fct_shift(
             relevel(
               as.factor(subfam),
               ref = 'ref')
           )
  )

#Need to add the transmission name to the subfam so that they are ordered correctly.
# Need to move 'ref' factor to the start of the subfam, then shift the factor levels to the 'left' (so now
# ref is at the end) ensuring that the 'ref' genes are at the end of the graph for each transmission type.

cir_plot$Geneid <- ifelse(
  cir_plot$Geneid %in% housekeeping_genes[1,],
  colnames(housekeeping_genes)[ match(cir_plot$Geneid, housekeeping_genes[1,]) ],
  cir_plot$Geneid
)

limits_cir <- c(0, 
                mround(max(cir_plot$tpm), 
                       base = 100))

@@

cir_plot <- mutate(cir_plot, 
                   Geneid = factor(Geneid, 
                                   levels = unique(cir_plot_order$Geneid)))

transmissionSep_cir_plot <- function(time_vec = '14h',
                                     log_it = TRUE,
                                     limits = limits_cir,
                                     point_size = 1,
                                     text_size = 10,
                                     threshold_min = 100,
                                     threshold_max = Inf){

  cir_plot_filter <- group_by(cir_plot, 
                              Geneid) %>% 
    filter(
      time == time_vec,
      !Geneid %in% cir_id | any(tpm >= threshold_min),
      !Geneid %in% cir_id | all(tpm <= threshold_max)
    )
  #Filter for pirs expressed above or below the thresholds, have the OR so that the ref genes aren't removed.
  # e.g. Either 'Not a cir' OR 'Any TPM is above the minimum'.

  # cir_plot_filter %>% group_by(Geneid) %>% summarise(n())
  if(log_it){
    cir_plot_filter$tpm <- log2(1+cir_plot_filter$tpm)
    limits = log2(1+limits)
    breaks <- round(seq.int(0, 
                            max(limits), 
                            length.out = round(max(limits)) + 1, 
                            digits = 0), 
                    digits = 0)
  }else{
    breaks <- seq.int(0, max(limits), length.out = 11)
  }

  cir_plot_filter$Geneid <- factor(
    str_remove(cir_plot_filter$Geneid, 
               pattern = 'PCHAS_'),
    levels = str_remove(levels(cir_plot_filter$Geneid), 
                        pattern = 'PCHAS_')
  )

  y <- ggplot(
    data = cir_plot_filter,
    mapping = aes(
      x = Geneid,
      y = tpm
    )
  ) +
    stat_summary(fun.y = median, 
                 fun.ymin = median, 
                 fun.ymax = median,
                 geom = "bar",
                 fill = 'lightblue',
                 width = 0.75, 
                 size = 0.25, 
                 position = 'dodge') +
    geom_point(stat = 'identity', 
               position = position_dodge(width = 0.75),
               size = point_size) +
    scale_y_continuous(breaks = breaks,
                       limits = limits) +
    facet_grid(.~subfam, 
               scales = "free_x",
               space = "free_x",
               labeller = labeller(subfam = function(string) {
                 str_remove_all(string, 
                                pattern = '(RMT|SBP)_')
               })) +
    theme_classic() +
    ylab('log(TPM + 1)')+
    ggtitle(paste(unique(cir_plot_filter$time)))

  if(grepl(time_vec, pattern = '20|23')) {
    y <- y + theme(
      axis.text.x = element_text(size = text_size,
                                 angle = 90,
                                 vjust = 0.5)
    )
  } else {
    y <- y + theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    ) 
  }

           #This bit is to change the colour, and text colour, of the facet boxes

           g <- ggplot_gtable(ggplot_build(y))

           # g$widths[5] = 2*g$widths[5]

           strip_t <- which(grepl('strip-t', g$layout$name))
           subfam_vec <- sort(unique(cir_plot_filter$subfam)) %>% str_remove_all(pattern = '(RMT|SBP)_')
           fills <- case_when(str_detect(subfam_vec, 
                                         pattern = 'a') ~ clade_col['ancestral'],
                              subfam_vec %in% names(clade_col) ~ clade_col[paste(subfam_vec)],
                              str_detect(subfam_vec, 
                                         pattern = 'ref') ~ 'white')
           k <- 1
           for (i in strip_t) {
             j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
             g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
             k <- k+1
           }

           #subfam_vec also corresponds to which panel the different clades are, so for the small ones we can use this to
           # manually make them a bit wider.

           short_panels <- which(str_detect(subfam_vec, pattern = 'a|S1'))

           for(i in short_panels){
             num <- g$layout$l[grep(paste0('panel-1-', i, '$'), g$layout$name)]
             g$widths[num] = 1.75*g$widths[num]
             #1.75 times the width of the right panels.
             # print(num)
           }

           if(global_save_var){
             ggsave2(
               paste0('Pchabaudi_side-by-side_barchart_facetbySubfam_',
                      time_vec,
                      '.png'),
               plot = g,
               dpi = 300,
               width = 28,
               height = 8,
               units = 'cm'
             )
           }

           grid.newpage()
           grid.draw(g)

}

lapply(
  sort(unique(cir_plot$time)),
  transmissionSep_cir_plot,
  threshold_min = 50,
  threshold_max = Inf,
  point_size = 1,
  text_size = 7.5
)
```

>>>>

cir_clade_info <- read_delim('../standard_files/Pchabaudi_pir_clades.txt',
            delim = '\t')

#Add chapl_aapl numbers for easier heatmap drawing later.
cir_info <- cir_info %>% 
  group_by(chapl_aapl_name) %>% 
  mutate(chapl_aapl_name_num = n_distinct(Geneid)) %>% 
  mutate(chapl_aapl_name_num = paste0(
                       chapl_aapl_name, 
                       ' (n = ',
                       chapl_aapl_name_num,
                       ')'
                     ))
```

@@

chapl_aapl_name_col <- setNames(
  c('#F0E442',
    '#56B4E9', #L1
    '#009E73',
    '#0072B2',
    '#E69F00', #S1
    '#D55E00',
    '#CC79A7',
    '#999999'),
  sort(
    unique(cir_info$chapl_aapl_name)
  )
)

chapl_aapl_name_col_num <- chapl_aapl_name_col
names(chapl_aapl_name_col_num) <- sapply(
  names(chapl_aapl_name_col), 
  function(clade_name) {
    paste0(
      clade_name, 
      ' (n = ',
      sum(cir_info$chapl_aapl_name == clade_name, 
          na.rm = TRUE),
      ')'
    )
  })
```

_pir_ genes selected by being expressed above 1 TPM in at least one sample.

@@

# Get order of pir genes by highest expressed genes in each sample.

combined_data_pir <- twentyfour_avgtpm$RMT$tpm %>% 
  filter(Geneid %in% cir_housekeeping_id) %>% 
  column_to_rownames(var = 'Geneid') %>% 
  filter_all(
      any_vars(. > 1)
  )

@@

@@

# Supp figure 9 - Pir expression of AAPLs and ChAPLs.

_pir_ genes selected by being expressed above 1 TPM in at least one sample.

```{r chapl_appl_heatmap_suppfig9}

# Get order of pir genes by highest expressed genes in each sample.

combined_data_pir <- twentyfour_avgtpm$RMT$tpm %>% 
  filter(Geneid %in% cir_id) %>% 
  mutate(chapl_aapl_name = cir_info$chapl_aapl_name_num[match(Geneid, cir_info$Geneid)]) %>% 
  filter(!str_detect(chapl_aapl_name, pattern = 'NA')) %>% 
  column_to_rownames(var = 'Geneid') %>% 
  filter_all(
      any_vars(. > 1)
  )

combined_data_ordered <- tibble(Geneid = rownames(combined_data_pir),
                                means = apply(combined_data_pir, 1, mean),
                                chapl_aapl_name = factor(
                                  combined_data_pir$chapl_aapl_name
                                )) %>% 
  arrange(chapl_aapl_name, 
          desc(means)) %>% 
  .$Geneid

#Order the heatmap data by the order defined above (chapl_aapl_name, then mean expression).
combined_data_pir_ht <- combined_data_pir[
  order(
    match(rownames(combined_data_pir),
          combined_data_ordered),
    na.last = NA #Remove NAs
  ),
]

#Change the colnames to the times
colnames(combined_data_pir_ht) <- str_extract(colnames(combined_data_pir_ht),
                                              pattern = '[[:digit:]]{2}h|chapl_aapl_name')

#Get the colour range
maxcol <- ceiling(
  max(
    log2(1+combined_data_pir_ht[!colnames(combined_data_pir_ht) %in% 'chapl_aapl_name'])
  ))

#Sub-family annotation
chapl_aapl_name_info <- combined_data_pir_ht$chapl_aapl_name
annotation_col <- c(chapl_aapl_name_col_num, 'ref' = 'black')

#chapl_aapl_name annotation

chapl_aapl_name_ha <- rowAnnotation(
  'sub-family' = anno_simple(
    chapl_aapl_name_info,
    col = annotation_col,
  ),
  annotation_name_gp = gpar(fontsize = 10)
)

#chapl_aapl_name name labels
num_chapl_aapl_name <- cir_info %>% 
  group_by(chapl_aapl_name) %>% 
  summarise(num = n_distinct(Gene.ID))

labels_chapl_aapl_name <- setNames(paste0(num_chapl_aapl_name$chapl_aapl_name, 
                                          ' (n = ', num_chapl_aapl_name$num, ')'),
                                   num_chapl_aapl_name$chapl_aapl_name)

ht <- Heatmap(
  log2(1+combined_data_pir_ht[,!colnames(combined_data_pir_ht) %in% 'chapl_aapl_name']),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  name = 'log(TPM+1)', 
  col = colorRamp2(c(0,
                     maxcol/4,
                     maxcol/2,
                     3*maxcol/4,
                     maxcol),
                   viridis(5)),
  na_col = 'white', 
  border = TRUE,
  border_gp = gpar(lwd = 0.1),
  column_names_gp = gpar(fontsize = 10),
  row_dend_width = unit(3, "cm"),
  width = unit(4.5, 'cm'),
  height = unit(7,'cm'),
  row_title = NULL,
  left_annotation = chapl_aapl_name_ha,
  row_split = 
    factor(combined_data_pir_ht$chapl_aapl_name,
              levels = ordering(combined_data_pir_ht$chapl_aapl_name,
                                order_vector = 'ChAPL')),
  heatmap_legend_param = list(
    labels_gp = gpar(fontsize = 10),
    title_gp = gpar(fontsize = 10,
                    fontface = 'bold'),
    at = c(0,
           maxcol/4,
           maxcol/2,
           3*maxcol/4,
           maxcol)
  )
)

chapl_aapl_name_lgd <- Legend(
  title = 'sub-family',
  at = factor(labels_chapl_aapl_name,
              ordering(labels_chapl_aapl_name,
                order_vector = 'ChAPL')),
  legend_gp = gpar(fill = annotation_col)
)

draw(ht, 
     row_title_side = 'left',
     column_title_side = 'bottom',
     heatmap_legend_side = 'right',
     annotation_legend_side = 'right',
     annotation_legend_list = list(chapl_aapl_name_lgd),
     merge_legend = TRUE)
```

```{r Supp_fig_AapL-ChaAPl_expression}

cir_plot <- twentyfour_rMT %>% 
  filter(Geneid %in% cir_id) %>%
  melt(id.vars = 'Geneid', 
       variable.name = 'sample', 
       value.name = 'tpm') %>% 
  mutate(
    stage = gsub(sample, pattern = '_[[:digit:]]$', replacement = ''),
    diff_count = str_split_fixed(sample, pattern = '_', n = 4)[,2],
    time = str_split_fixed(sample, pattern = '_', n = 3)[,1],
    subfam = cir_clade_info$clade[match(Geneid, cir_clade_info$Geneid)],
    chapl_aapl = factor(cir_info$chapl_aapl_name[match(Geneid, 
                                                       cir_info$Geneid)],
                        levels = sort(unique(cir_info$chapl_aapl_name[match(Geneid, 
                                                                            cir_info$Geneid)])))) %>% 
  filter(!(time %in% c('sch', 
                       'Unknown')),
         !is.na(chapl_aapl) ) %>% 
  group_by(Geneid) %>% 
    arrange(desc(tpm))

@@

  cir_plot <- mutate(cir_plot, 
                     Geneid = factor(Geneid, 
                                     levels = unique(cir_plot_order$Geneid)))
  cir_plot_chapl_aapl_supp_info <- cir_plot

  max_aapl_chapl <- ceiling(max(log2(cir_plot$tpm+1)))

  breaks = seq(0, max_aapl_chapl, by = 1)
    limits = c(0, max_aapl_chapl)
    point_size = 0.25

  lapply(
    sort(unique(cir_plot$time)),
    function(time_vec){

    cir_plot <- filter(cir_plot, 
                       time == time_vec)

      cir_plot$tpm <- log2(1+cir_plot$tpm)

    cir_plot$Geneid <- factor(
      str_remove(cir_plot$Geneid, 
                 pattern = 'PCHAS_'),
      levels = str_remove(levels(cir_plot$Geneid), 
                          pattern = 'PCHAS_')
    )

    y <- ggplot(
      data = cir_plot,
      mapping = aes(
        x = Geneid,
        y = tpm
      )
    ) +
      stat_summary(fun.y = median, 
                   fun.ymin = median, 
                   fun.ymax = median,
                   geom = "bar", 
                   width = 0.75, 
                   size = 0.25, 
                   position = 'dodge',
                   fill = 'lightblue') +
      geom_point(stat = 'identity', 
                 position = position_dodge(width = 0.75),
                 size = point_size) +
      scale_y_continuous(breaks = breaks,
                         limits = limits) +
      scale_fill_manual(values = c('skyblue','red')) +
      ylab('log(TPM +1)') +
      facet_grid(.~chapl_aapl, 
                 scales = "free_x",
                 space = "free_x") +
      theme_classic() +
      theme(
        axis.text.x = element_text(angle = 90,
                                   size = 5)
      ) +
      ggtitle(paste(unique(cir_plot$time)))

    if(global_save_var){
      ggsave2(
        paste0('Pchabaudi_sibe-by-side_chapl_aapl_',
               time_vec,
               '.png'),
        plot = y,
        dpi = 300,
        width = 25,
        height = 10,
        units = 'cm'
      )
    }

    return(y)
    })
```
